---
title: "Interactive Visualization with the songR Shiny App"
author: "Raban Heller"
date: "`r Sys.Date()`"
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/r-heller/songR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/r-heller/songR/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/r-heller/songR/actions/workflows/pkgdown.yaml/badge.svg)](https://r-heller.github.io/songR/)
[![CRAN status](https://www.r-pkg.org/badges/version/songR)](https://CRAN.R-project.org/package=songR)
[![Codecov test coverage](https://codecov.io/gh/r-heller/songR/branch/main/graph/badge.svg)](https://app.codecov.io/gh/r-heller/songR?branch=main)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/songR)](https://cran.r-project.org/package=songR)
[![CRAN downloads total](https://cranlogs.r-pkg.org/badges/grand-total/songR)](https://cran.r-project.org/package=songR)
[![License: MIT](https://img.shields.io/badge/license-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

songR includes a built-in Shiny application for interactive exploration
of SONG embeddings. The app lets you:

- Upload CSV/RDS data or use built-in datasets
- Adjust SONG hyperparameters (epochs, epsilon, spread_factor, k)
- Run SONG, t-SNE, and UMAP side-by-side
- Incrementally add data and observe embedding stability
- Color points by any variable (discrete or continuous)
- Toggle **dark mode** for comfortable viewing (preference persisted via
  localStorage)
- Export embeddings and plots

All plots use the **viridis plasma** color scale for perceptually
uniform, colorblind-friendly visualization.

## Launching the App

```{r launch}
library(songR)

# Launch with default settings
run_songR_app()
```

The app opens in your default browser.

## Using Built-in Data

The app ships with `songR_blobs`, an 8-cluster synthetic dataset in 20
dimensions. Select it from the "Data" dropdown to get started immediately.

## Uploading Your Own Data

1. Click **Upload Data** in the sidebar
2. Select a CSV file (first columns = features, optional last column = labels)
3. Or upload an RDS file containing a list with `$data` (matrix) and
   optional `$labels` (factor/character)

## Hyperparameter Controls

The sidebar provides sliders for all key SONG parameters:

| Control | Maps to | Effect |
|---------|---------|--------|
| Epochs | `epochs` | More = better quality, slower |
| Epsilon | `epsilon` | Edge decay rate (0-1) |
| Spread Factor | `spread_factor` | Growth threshold for new CVs |
| k | `k` | Neighborhood size |
| Dispersion | `dispersion` | Toggle UMAP refinement |

## Incremental Mode

The app supports SONG's incremental workflow:

1. Load initial data and click **Run SONG**
2. Upload additional data via **Add Increment**
3. Click **Update Model** — the existing embedding is preserved
4. Compare with **Reinitialize** to see the stability difference

## Dark Mode

Click the moon/sun toggle button in the navbar to switch between light
and dark mode. The preference is saved in your browser's localStorage
and restored on the next visit.

Dark mode adjusts:

- Background, text, and border colors for all UI panels
- Plot backgrounds and axis label colors
- Sidebar, navbar, and well panel styling

## Exporting Results

- **Download Embedding**: Saves the 2D coordinates as CSV
- **Download Plot**: Exports the current visualization as PNG/PDF
- **Download Model**: Saves the full `song_model` object as RDS

## Example Workflow

```{r workflow}
library(songR)

# Programmatic equivalent of the Shiny workflow:

# 1. Load data
data(songR_blobs)
X <- songR_blobs$data
labels <- songR_blobs$labels

# 2. Initial SONG
model <- song(X[1:500, ], epochs = 20L, seed = 42, verbose = FALSE)
plot(model, color_by = labels[1:500])

# 3. Incremental update
model <- update(model, X[501:1000, ], epochs = 20L, verbose = FALSE)
emb <- predict(model, newdata = X[1:1000, ])

# 4. Visualize
plot(emb[, 1], emb[, 2], pch = 16, cex = 0.5,
     col = rainbow(8)[as.integer(labels[1:1000])],
     main = "After Incremental Update",
     xlab = "SONG 1", ylab = "SONG 2", bty = "n")
```

## Session Info

```{r sessioninfo, eval = TRUE}
sessionInfo()
```
